Syntax and Semantics of Abstract Binding Trees

### 2015-11-12 paper.tex
[paper] fix chktex warnings

Abstract binding trees (abts) are a generalization of abstract syntax trees
where operators may express variable binding structure as part of their
arities.  Originally formulated by Peter Aczel \cite{aczel:1978}, unisorted
abts have been deployed successfully as the uniform syntactic framework for
several implementations of Constructive Type Theory, including Nuprl
\cite{constable:1986}, MetaPRL \cite{hickey:2003} and JonPRL
\cite{jonprl:2015}.

In \emph{Practical Foundations for Programming Languages} \cite{harper:2016},
Robert Harper develops the multi-sorted version of abstract binding trees and
proposes an extension to include families of operators indexed by
\emph{symbols}, which are, unlike variables, subject to only
distinctness-preserving renaming and not substitution; furthermore, symbols do
not appear in the syntax of abts, and are only introduced as parameters to
operators. This extension to support symbols is essential for a
correct treatment of programming languages with open sums (e.g. ML's
\texttt{exn} type), as well as assignable references.

@@

The contribution of this paper is the development of the syntax and semantics
of multi-sorted abts, an extension of second order universal
algebra to support symbol-indexed families of operators. Additionally, we have
developed the categorical semantics for abts formally in Constructive
Type Theory using the Agda proof assistant \cite{norell:2007}.

@@

Having specified an approximation of the grammar of telescopes and sequents in
the abt logical framework, we can proceed to define proper wellformedness via
\emph{inductive refinement} \cite{harper:2016}. The basic idea is to introduce a new form of
(meta)-judgment $\IsWf{\Omega}{\Upsilon}{\Gamma}{M}{\tau}$ which expresses the
extrinsic wellformedness properties we wish to verify, presupposing
$\IsAbt{\Omega}{\Upsilon}{\Gamma}{M}{\tau}$. Additionally, we introduce an
analogous judgment on bound terms, $\IsWf{\Omega}{\Upsilon}{\Gamma}{E}{v}$
presupposing $\IsAbt{\Omega}{\Upsilon}{\Gamma}{E}{v}$, defined uniformly as
follows:
\[
  \infer{
    \IsWf{\Omega}{\Upsilon}{\Gamma}{
      \MkBTm{\vec{u}}{\vec{x}}{M}
    }{
      \MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}
    }
  }{
    \IsWf{\Omega}{\Upsilon,\vec{u}:\vec{\sigma}}{\Gamma,\vec{x}:\vec{\tau}}{M}{\tau}
  }
\]

>>>>

Abstract binding trees (abts) are a generalization of abstract syntax trees
where operators may express variable binding structure as part of their
arities.  Originally formulated by Peter Aczel~\cite{aczel:1978}, unisorted
abts have been deployed successfully as the uniform syntactic framework for
several implementations of Constructive Type Theory, including Nuprl
\cite{constable:1986}, MetaPRL~\cite{hickey:2003} and JonPRL
\cite{jonprl:2015}.

In \emph{Practical Foundations for Programming Languages}~\cite{harper:2016},
Robert Harper develops the multi-sorted version of abstract binding trees and
proposes an extension to include families of operators indexed by
\emph{symbols}, which are, unlike variables, subject to only
distinctness-preserving renaming and not substitution; furthermore, symbols do
not appear in the syntax of abts, and are only introduced as parameters to
operators. This extension to support symbols is essential for a
correct treatment of programming languages with open sums (e.g. ML's
\texttt{exn} type), as well as assignable references.

@@

The contribution of this paper is the development of the syntax and semantics
of multi-sorted abts, an extension of second order universal
algebra to support symbol-indexed families of operators. Additionally, we have
developed the categorical semantics for abts formally in Constructive
Type Theory using the Agda proof assistant~\cite{norell:2007}.

@@

Having specified an approximation of the grammar of telescopes and sequents in
the abt logical framework, we can proceed to define proper wellformedness via
\emph{inductive refinement}~\cite{harper:2016}. The basic idea is to introduce a new form of
(meta)-judgment $\IsWf{\Omega}{\Upsilon}{\Gamma}{M}{\tau}$ which expresses the
extrinsic wellformedness properties we wish to verify, presupposing
$\IsAbt{\Omega}{\Upsilon}{\Gamma}{M}{\tau}$. Additionally, we introduce an
analogous judgment on bound terms, $\IsWf{\Omega}{\Upsilon}{\Gamma}{E}{v}$
presupposing $\IsAbt{\Omega}{\Upsilon}{\Gamma}{E}{v}$, defined uniformly as
follows:
\[
  \infer{
    \IsWf{\Omega}{\Upsilon}{\Gamma}{
      \MkBTm{\vec{u}}{\vec{x}}{M}
    }{
      \MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}
    }
  }{
    \IsWf{\Omega}{\Upsilon,\vec{u}:\vec{\sigma}}{\Gamma,\vec{x}:\vec{\tau}}{M}{\tau}
  }
\]
